final-project

### 2024-11-13 final-project.Rmd
ph assumption

```{r}
# column rename
seer_name <- seer1 %>%
  dplyr::rename(
  age = Age.recode.with..1.year.olds,
  sex = Sex,
  year_diagnosis = Year.of.diagnosis,
  race = Race.ethnicity,
  marriage_status = Marital.status.at.diagnosis,
  tnm2015 = Derived.AJCC.T..7th.ed..2010.2015.,
  tnm2017 = Derived.SEER.Combined.T..2016.2017.,
  tnm2018 = Derived.EOD.2018.T..2018..,
  subtype = Breast.Subtype..2010..,
  grade2017 = Grade.Recode..thru.2017.,
  grade2018 = Derived.Summary.Grade.2018..2018..,
  n2015 = Derived.AJCC.N..7th.ed..2010.2015.,
  n2017 = Derived.SEER.Combined.N..2016.2017.,
  n2018 = Derived.EOD.2018.N..2018..,
  m2015 = Derived.AJCC.M..7th.ed..2010.2015.,
  m2017 = Derived.SEER.Combined.M..2016.2017.,
  m2018 = Derived.EOD.2018.M..2018..,
  radiation = Radiation.recode,
  chemotherapy = Chemotherapy.recode..yes..no.unk.,
  surgery = RX.Summ..Surg.Prim.Site..1998..,
  survival_months = Survival.months,
  survival_status = Vital.status.recode..study.cutoff.used.,
  survival_cause_cancer = SEER.cause.specific.death.classification,
  survival_cause_other = SEER.other.cause.of.death.classification,
  idc_type = Histologic.Type.ICD.O.3
)

# filter
seer_filter <- seer_name %>%
  filter(
    age != "00 years",
    age != "15-19 years",
    sex == "Female",
    survival_months >= 1,
    survival_cause_cancer != "Dead (missing/unknown COD)",
    survival_cause_other != "Dead (attributable to causes other than this cancer dx)",
    grade2017 != "B-cell; pre-B; B-precursor",
    grade2017 != "T-cell",
    idc_type %in% seq(8500, 8549)
  )

# clean up the variables
seer_clean <- seer_filter %>%
  mutate(
    # assign race as White, Black, or Other
    race = case_when(
      race == "White" ~ "White",
      race == "Black" ~ "Black",
      TRUE ~ "Other"
    ),
    # assign marriage_status as Yes or No
    marriage_status = case_when(
      marriage_status == "Married (including common law)" ~ "Yes",
      TRUE ~ "No"
    ),

@@

    # simplify the radiation variable
    radiation = case_when(
      str_detect(radiation, "None|Refused|unknown") ~ "No/Unknown",
      TRUE ~ "Yes"
    ),
    surgery = case_when(
      surgery == "0" ~ "No/Unknown",
      surgery == "99" ~ "No/Unknown",
      TRUE ~ "Yes"
    )
    )

# there are several variables that need to be categorized for analysis and/filtered
seer_cat <- seer_clean %>%
   filter(
    !is.na(subtype),
    #!is.na(grade),
    #n_stage != "NX"
  ) %>%
  mutate(agecat = case_when(
    age %in% c("20-24 years", "25-29 years", "30-34 years", "35-39 years", "40-44 years", "45-49 years") ~ 0,
    age %in% c("50-54 years", "55-59 years", "60-64 years", "65-69 years") ~ 1,
    age %in% c("70-74 years", "75-79 years", "80-84 years", "85+ years") ~ 2,
    TRUE ~ NA_real_  # Keeps other values as NA if they don’t match any of the above groups
  )) %>%
  mutate(gradecat = case_when(
    grade %in% c(1, 2) ~ "I/II",
    grade %in% c(3, 4) ~ "III/IV",
    TRUE ~ NA_character_
    )) %>%
  # the survival_status column is what we use for event/censoring
  # recode Dead as 1 and Alive (censored) as 0
  mutate(status_cat = case_when(
    survival_status == "Dead" ~ 1,
    survival_status == "Alive" ~ 0
  ))

# further filter and select only variables used in analysis
seer <- seer_cat %>%
  select(
    age = agecat,
    sex,
    year_diagnosis,
    race,
    marriage_status,
    subtype,
    n_stage,
    m_stage,
    grade = gradecat,
    radiation,
    chemotherapy,
    surgery,
    survival_months,
    survival_status = status_cat,

@@

```{r}
# check the proportional hazards assumption for each covariate

## age
ggsurvplot(survfit(Surv(time = survival_months, event = survival_status) ~ age), data = seer, fun="cloglog")

@@

>>>>

```{r}
# column rename
seer_name <- seer1 %>%
  dplyr::rename(
  age = Age.recode.with..1.year.olds,
  sex = Sex,
  year_diagnosis = Year.of.diagnosis,
  race = Race.ethnicity,
  marriage_status = Marital.status.at.diagnosis,
  tnm2015 = Derived.AJCC.T..7th.ed..2010.2015.,
  tnm2017 = Derived.SEER.Combined.T..2016.2017.,
  tnm2018 = Derived.EOD.2018.T..2018..,
  subtype = Breast.Subtype..2010..,
  grade2017 = Grade.Recode..thru.2017.,
  grade2018 = Derived.Summary.Grade.2018..2018..,
  n2015 = Derived.AJCC.N..7th.ed..2010.2015.,
  n2017 = Derived.SEER.Combined.N..2016.2017.,
  n2018 = Derived.EOD.2018.N..2018..,
  m2015 = Derived.AJCC.M..7th.ed..2010.2015.,
  m2017 = Derived.SEER.Combined.M..2016.2017.,
  m2018 = Derived.EOD.2018.M..2018..,
  radiation = Radiation.recode,
  chemotherapy = Chemotherapy.recode..yes..no.unk.,
  surgery = RX.Summ..Surg.Prim.Site..1998..,
  survival_months = Survival.months,
  survival_status = Vital.status.recode..study.cutoff.used.,
  survival_cause_cancer = SEER.cause.specific.death.classification,
  survival_cause_other = SEER.other.cause.of.death.classification,
  idc_type = Histologic.Type.ICD.O.3
)
```

```{r}
# filter
seer_filter <- seer_name %>%
  filter(
    age != "00 years",
    age != "15-19 years",
    sex == "Female",
    survival_months >= 1,
    survival_cause_cancer != "Dead (missing/unknown COD)",
    survival_cause_other != "Dead (attributable to causes other than this cancer dx)",
    grade2017 != "B-cell; pre-B; B-precursor",
    grade2017 != "T-cell",
    idc_type %in% seq(8500, 8549)
  )
```

```{r}
# clean up the variables
seer_clean <- seer_filter %>%
  mutate(
    # assign race as White, Black, or Other
    race = case_when(
      race == "White" ~ "White",
      race == "Black" ~ "Black",
      TRUE ~ "Other"
    ),
    # assign marriage_status as Yes or No
    marriage_status = case_when(
      marriage_status == "Married (including common law)" ~ "Yes",
      TRUE ~ "No"
    ),

@@

    # simplify the radiation variable
    radiation = case_when(
      str_detect(radiation, "None|Refused|unknown") ~ "No/Unknown",
      TRUE ~ "Yes"
    ),
    surgery = case_when(
      surgery == "0" ~ "No/Unknown",
      surgery == "99" ~ "No/Unknown",
      TRUE ~ "Yes"
    )
    )
```

```{r}
# there are several variables that need to be categorized for analysis and/filtered
seer_cat <- seer_clean %>%
   filter(
    !is.na(subtype),
    #!is.na(grade),
    #n_stage != "NX"
  ) %>%
  mutate(agecat = case_when(
    age %in% c("20-24 years", "25-29 years", "30-34 years", "35-39 years", "40-44 years", "45-49 years") ~ 0,
    age %in% c("50-54 years", "55-59 years", "60-64 years", "65-69 years") ~ 1,
    age %in% c("70-74 years", "75-79 years", "80-84 years", "85+ years") ~ 2,
    TRUE ~ NA_real_  # Keeps other values as NA if they don’t match any of the above groups
  )) %>%
  mutate(agecat0 = case_when(
    agecat == 0 ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(agecat1 = case_when(
    agecat == 1 ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(agecat2 = case_when(
    agecat == 2 ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(racecat = case_when(
    race == "White" ~ 0,
    race == "Black" ~ 1,
    race == "Other" ~ 2,
    TRUE ~ NA_integer_
  )) %>%
  mutate(racecat0 = case_when(
    race == "White" ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(racecat1 = case_when(
    race == "Black" ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(racecat2 = case_when(
    race == "Other" ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(marriage_status = case_when(
    marriage_status == "Yes" ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(subtypecat = case_when(
    subtype == "Luminal A" ~ 0,
    subtype == "Luminal B" ~ 1,
    subtype == "Triple Negative" ~ 2,
    subtype == "HER2 Positive" ~ 3,
    TRUE ~ NA_integer_
  )) %>%
  mutate(subtype0 = case_when(
    subtypecat == 0 ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(subtype1 = case_when(
    subtypecat == 1 ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(subtype2 = case_when(
    subtypecat == 2 ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(subtype3 = case_when(
    subtypecat == 3 ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(gradecat = case_when(
    grade %in% c(1, 2) ~ 0,
    grade %in% c(3, 4) ~ 1,
    TRUE ~ NA_integer_
    )) %>%
  mutate(n_stage = case_when(
    n_stage %in% c("N0", "N1") ~ 0,
    n_stage %in% c("N2", "N3") ~ 1,
    TRUE ~ NA_integer_
    )) %>%
  mutate(m_stage = case_when(
    m_stage == "M0" ~ 0,
    m_stage == "M1" ~ 1,
    TRUE ~ NA_integer_
    )) %>%
  mutate(radiation = case_when(
    radiation == "No/Unknown" ~ 0,
    radiation == "Yes" ~ 1,
    TRUE ~ NA_integer_
    )) %>%
  mutate(chemotherapy = case_when(
    chemotherapy == "No/Unknown" ~ 0,
    chemotherapy == "Yes" ~ 1,
    TRUE ~ NA_integer_
    )) %>%
  mutate(surgery = case_when(
    surgery == "No/Unknown" ~ 0,
    surgery == "Yes" ~ 1,
    TRUE ~ NA_integer_
    )) %>%
  # the survival_status column is what we use for event/censoring
  # recode Dead as 1 and Alive (censored) as 0
  mutate(status_cat = case_when(
    survival_status == "Dead" ~ 1,
    survival_status == "Alive" ~ 0
  ))
```

```{r}
# further filter and select only variables used in analysis
seer <- seer_cat %>%
  select(
    agecat,
    agecat0,
    agecat1,
    agecat2,
    racecat,
    racecat0,
    racecat1,
    racecat2,
    marriage_status,
    subtype = subtypecat,
    subtype0,
    subtype1,
    subtype2,
    subtype3,
    n_stage,
    m_stage,
    grade = gradecat,
    radiation,
    chemotherapy,
    surgery,
    survival_months,
    survival_status = status_cat,

@@

```{r}
# check the proportional hazards assumption for each covariate with (1) a log log plot

## subtype
ggsurvplot(survfit(Surv(time = survival_months, event = survival_status) ~ subtype), data = seer, fun="cloglog")

## agecat
ggsurvplot(survfit(Surv(time = survival_months, event = survival_status) ~ agecat), data = seer, fun="cloglog")

@@

```{r}
# create a dataset in counting process format
seer.cpfull <- survSplit(seer, cut=seer$survival_months[seer$survival_status==1],
                     end="survival_months",event="survival_status", start="start")
# remove NA because the model will do this anyways
seer.cp <- na.omit(seer.cpfull)
```

```{r}
# check PH assumption for race, n_stage, chemotherapy, surgery with statistical approach

# create time-varying covariates for race by adding adding a product term with the ln of time
seer.cp$logtrace1=seer.cp$race[seer.cp$race == "White"]*log(seer.cp$survival_months)

seer.cp$logtrace <- ifelse(seer.cp$race == "White", log(seer.cp$survival_months), NA)

```

```{r}
model <- coxph(
  Surv(start, survival_months, survival_status) ~ relevel(factor(subtype), ref = '0') +
    relevel(factor(agecat), ref = '0') + 
    relevel(factor(racecat), ref = '0') +
    relevel(factor(marriage_status),ref='1') + 
    relevel(factor(n_stage),ref='0') + 
    relevel(factor(m_stage),ref='0') +
    relevel(factor(grade), ref='0') +
    relevel(factor(radiation), ref='1') + 
    relevel(factor(chemotherapy), ref='1') +
    relevel(factor(surgery), ref='1'),
  data = seer.cp, 
  method = "breslow")

summary(model)

# Test proportional hazards assumption with Schoenfeld residuals
cox.zph(model)
```
